---
title: "10 Insights from New Mexico School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Insights from New Mexico School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

```{r load-packages}
library(nmschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores New Mexico's public school enrollment data, surfacing key trends and demographic patterns across 10 years of data (2016-2025).

---

## 1. New Mexico's enrollment has been declining

Like many western states, New Mexico's K-12 enrollment has been steadily declining over the past several years, with COVID accelerating the trend.

```{r statewide-trend}
enr <- fetch_enr_multi(2019:2025)

state_totals <- enr |>
  filter(is_state, subgroup == "total", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#C41230") +
  geom_point(size = 3, color = "#C41230") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "New Mexico Public School Enrollment (2019-2025)",
    subtitle = "Steady decline accelerated by COVID-19",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Albuquerque dominates but is shrinking fastest

Albuquerque Public Schools serves roughly one-quarter of all New Mexico students, but the district has been losing enrollment faster than the state average.

```{r top-districts}
top_districts <- enr |>
  filter(is_district, subgroup == "total", grade_level == "TOTAL",
         end_year == 2025) |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

top_districts
```

```{r top-districts-chart}
top_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "New Mexico's Largest School Districts (2025)",
    subtitle = "APS serves about 1 in 4 NM students",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 3. A Hispanic majority state

New Mexico is one of the most diverse states in the nation. Hispanic students make up over 60% of enrollment, with significant Native American representation.

```{r demographics}
enr_2025 <- fetch_enr(2025)

demographics <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("hispanic", "white", "native_american", "black", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

demographics
```

```{r demographics-chart}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "New Mexico Student Demographics (2025)",
    subtitle = "Hispanic students are the clear majority",
    x = "Number of Students",
    y = NULL
  )
```

---

## 4. The Native American education landscape

New Mexico has the third-highest Native American student population in the country, concentrated in districts near the Navajo Nation, Zuni, and numerous Pueblos.

```{r native-american}
native_am <- enr_2025 |>
  filter(is_district, subgroup == "native_american", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

native_am
```

---

## 5. COVID crushed kindergarten

New Mexico kindergarten enrollment dropped significantly during COVID and hasn't fully recovered, signaling smaller cohorts for years to come.

```{r covid-k}
covid_grades <- enr |>
  filter(is_state, subgroup == "total",
         grade_level %in% c("K", "01", "06", "09"),
         end_year %in% 2019:2025) |>
  select(end_year, grade_level, n_students) |>
  pivot_wider(names_from = grade_level, values_from = n_students)

covid_grades
```

```{r growth-chart}
enr |>
  filter(is_state, subgroup == "total",
         grade_level %in% c("K", "01", "06", "09")) |>
  ggplot(aes(x = end_year, y = n_students, color = grade_level)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Enrollment by Grade Level Over Time",
    subtitle = "Kindergarten shows steepest COVID-era decline",
    x = "School Year",
    y = "Students",
    color = "Grade"
  )
```

---

## 6. Albuquerque vs Las Cruces: Two cities diverging

Albuquerque and Las Cruces are New Mexico's two largest districts, but their enrollment trajectories have differed.

```{r abq-lc}
abq_lc <- enr |>
  filter(is_district, subgroup == "total", grade_level == "TOTAL",
         grepl("Albuquerque|Las Cruces", district_name)) |>
  select(end_year, district_name, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students)

abq_lc
```

```{r abq-lc-chart}
enr |>
  filter(is_district, subgroup == "total", grade_level == "TOTAL",
         grepl("Albuquerque|Las Cruces", district_name)) |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Albuquerque vs Las Cruces Enrollment",
    subtitle = "The state's two largest districts",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## Summary

New Mexico's school enrollment data reveals:

- **Declining enrollment**: The state has lost students steadily since 2019
- **Hispanic majority**: Over 60% of students are Hispanic
- **Native American presence**: Third-highest tribal enrollment in the nation
- **Urban concentration**: APS serves 1 in 4 students statewide
- **COVID impact**: Kindergarten hit hardest, rippling through cohorts

These patterns shape school funding debates and facility planning across the Land of Enchantment.

---

*Data sourced from the New Mexico Public Education Department [STARS System](https://web.ped.nm.gov/bureaus/information-technology/stars/).*
